\section{Vertical Inserters}

We define vertical inserters, which are particular marked 2-limits~\cite[I.7.9.1.(iii)]{GrayMarkedLimits}.
In this section, $ℬ$ denotes an arbitrary 2-category.
\begin{definition}
  We define a 2-category $\VInsDiag(ℬ) $ of vertical inserter diagrams (abbreviated as
  vins-diagrams) which is a wide, locally full sub 2-category 
  of the 2-category of functors from 
  $\begin{tikzcd}
   ⋅ \doublerightarrow{r}{}{}& ⋅ \ar[r] & ⋅  
  \end{tikzcd}
  $
  to $ℬ$, lax natural transformations and modifications~\cite{twodimcat}.
   An object is given by a diagram of three 1-cells 
   $
   \begin{tikzcd}
   A \doublerightarrow{r}{F}{G} & B \ar[r,"p"] & C.
   \end{tikzcd}
   $ such that $p∘ F = p∘  G$.
   A 1-cell between $(F,G,p)$ and $(F',G',p')$ is given 
   by three 1-cells $(A \xrightarrow{H_A} A', B\xrightarrow{H_B} B', C\xrightarrow{H_C} C')$ such that 
   $H_C 	
   ∘
   p = p' 	
   ∘
    H_B$ and $H_B ∘ G = G' ∘ H_A$, together with
   a 2-cell 
   $  F' H_A \xrightarrow{h_F}  H_BF$
   as summarised in the following diagram, such that 
   $p' ⋅ h_F$ is the identity 2-cell on $p' ∘ H_B ∘ F = p' ∘ F' ∘ H_A $.
   \[
   % YADE DIAGRAM diagrams/vins-diag-1-cell.json
% GENERATED LATEX
\input{../diagrams/vins-diag-1-cell.tex}
% END OF GENERATED LATEX
\]
% En termes de string diagram (bottom up, left right)
% h_F 
%   F  H_B
%   H_A F'

% h_G 
%   H_A G'
%   G H_B

%h_p est une egalite
A 2-cell between $(H_A, H_B, H_C, h_F)$ and
$(H_A', H_B', H_C', h_F')$ is given by three 2-cells
$(
  H_A \xrightarrow{h_A}H_A,
  H_B \xrightarrow{h_B}H_B,
  H_C \xrightarrow{h_C}H_C
)$ such that the following equalities hold:
\[
% YADE DIAGRAM diagrams/vins-2cell.json  
% GENERATED LATEX
\input{../diagrams/vins-2cell.tex}
% END OF GENERATED LATEX
\]

\end{definition}
\begin{definition}
  There is a canonical 2-functor $Δ ∶ ℬ → \VInsDiag(ℬ)$ mapping an object  
   $b$ to the constant diagram which is pointwise $b$~\cite[I.7.9.1.(iii)]{GrayMarkedLimits}.
  We say that $ℬ$ has vertical inserters if $Δ$ has a right (2-)adjoint.
  In this case, 
  the \emph{vertical inserter} of a vins-diagram $(F,G,p)$ is defined as
  its image by the right adjoint.
\end{definition}
\begin{example}
  A vertical inserter of a vins-diagram as above in $\Cat$ is the category 
  whose objects are morphisms $ F a → G a $ such that their image by $p$
  is the identity morphism on $p F a = p G a$, and morphisms are the compatible
  morphisms in $A$.
\end{example}
Vertical inserters have a universal property (they are limits)
that we now spell out.
% \begin{remark}
%   The vertical inserter of a vins-diagram $
%   \begin{tikzcd}
%   A \doublerightarrow{r}{F}{G} & B \ar[r,"p"] & C.
%   \end{tikzcd}$ comes equipped with
% \end{remark}

\begin{definition}
  A cone of a vins-diagram $   \begin{tikzcd}
       A \doublerightarrow{r}{F}{G} & B \ar[r,"p"] & C.
       \end{tikzcd}$ over a 0-cell $X$ 
  consists of the following data:
   \begin{itemize}
    \item a 1-cell $X \xrightarrow{x}A$ such that
     $p∘ F∘ x = p ∘ G ∘ x$;
    \item a 2-cell  $ F · x \xrightarrow{α} G · x$ such that 
     $p ⋅ α$ is the 2-cell
     on $p∘ F∘ x = p ∘ G ∘ x$.
   \end{itemize}
   A morphism between two cones $(x,α)$ and $(x',α')$ over $X$ 
   consists in a natural transformation $γ ∶ x → x'$ such that
   the following equality holds.
   \begin{equation}
    \label{eq:mor-cones}
    % YADE DIAGRAM diagrams/ins-2universal-prop.json
% GENERATED LATEX
\input{../diagrams/ins-2universal-prop.tex}
% END OF GENERATED LATEX
  \end{equation}
\end{definition} 
\begin{lemma}
  The vertical inserter $I$ of a 2-diagram
  $   \begin{tikzcd}
    A \doublerightarrow{r}{F}{G} & B \ar[r,"p"] & C
    \end{tikzcd}$ comes with a cone $(i,κ)$ over it which is universal in the following sense.
    \begin{itemize}
        \item 
      Any cone $ F · x \xrightarrow{α} G · x$  is of the shape 
      $κ ⋅ \universalins{\alpha}$ for a unique 
      $\universalins{\alpha}$, as summarised in the diagram below.
      \[
    % YADE DIAGRAM diagrams/ins-universal-prop.json
    % GENERATED LATEX
    \input{../diagrams/ins-universal-prop.tex}
    % END OF GENERATED LATEX
    \]
    \item
      Any cone morphism
      $x \xrightarrow{γ} x'$
      between  $ F · x \xrightarrow{α} G · x$ and 
      $ F · x' \xrightarrow{α'} G · x'$ is of the shape 
      $\kappa ⋅ \universalins{\gamma}$ 
      for a unique $\universalins{\gamma}∶ \universalins{\alpha} → \universalins{\alpha'}$, as summarised in the diagram below.
    \[ 
      % YADE DIAGRAM diagrams/ins-2universal-last.json
    % GENERATED LATEX
    \input{../diagrams/ins-2universal-last.tex}
    % END OF GENERATED LATEX
      \]
    \end{itemize}
\end{lemma}
 
%%% Local Variables:
%%% TeX-master: "../main.tex"
%%% End:
